Extreme precip and heat waves: empirical-statistical downscaling and analysis

### 2016-09-22 ClimaTrans/climatrans.Rmd
ClimaTrans updates

```{r,echo=TRUE}
library(esd)
library(ncdf)
#rm(list=ls())
## Information about the system and session
```

@@

```{r, eval=FALSE}
hotsummerdays <- function (x, y = NULL, dse = NULL, it = "jja", threshold = 30, 
    verbose = FALSE, plot = TRUE, nmin = 90, new = TRUE, ...) 
{
    if (verbose) 
        print("mildwinterdays")
    stopifnot(inherits(x, "station"))
    if (is.null(y)) 
        y <- x
    djf <- subset(x, it = it)
    djfy <- subset(y, it = it)
    nwd1 <- annual(djfy, FUN = "count", threshold = threshold, 
        nmin = nmin)
    mwd1 <- annual(djf, FUN = "mean", nmin = nmin)
    cal <- data.frame(x = c(coredata(mwd1)), y = c(coredata(nwd1)))
    dfit <- glm(y ~ x, family = "poisson", data = cal)
    if (plot) {
        if (new) 
            dev.new()
        par(bty = "n")
        plot(cal, pch = 19, ylim = c(0, 90), xlab = expression(paste("mean temperature ", 
            (degree * C))), ylab = "number of hot days", main = loc(x))
        pre <- data.frame(x = seq(min(cal$x, na.rm = TRUE) - 
            1, max(cal$x, na.rm = TRUE) + 5, by = 0.1))
        lines(pre$x, exp(predict(dfit, newdata = pre)), col = rgb(1, 
            0, 0, 0.3), lwd = 3)
        if (new) 
            dev.new()
        djf.sd <- sd(coredata(djf), na.rm = TRUE)
        qqnorm(coredata(djf))
        qqline(coredata(coredata(djf)), col = "red")
        grid()
    }
    if (is.null(dse)) 
        dse <- DSensemble.t2m(x, biascorrect = TRUE, verbose = verbose, 
            plot = plot)
    djf.dse <- subset(dse, it = "djf")
    index(djf.dse) <- year(djf.dse)
    ovl <- window(djf.dse, start = year(start(x)), end = year(end(x)))
    djf.dse <- djf.dse - mean(coredata(ovl), na.rm = TRUE) + 
        mean(coredata(mwd1), na.rm = TRUE)
    q1 <- data.frame(x = apply(coredata(djf.dse), 1, quantile, 
        probs = 0.05, na.rm = TRUE))
    q2 <- data.frame(x = apply(coredata(djf.dse), 1, quantile, 
        probs = 0.95, na.rm = TRUE))
    qm <- data.frame(x = apply(coredata(djf.dse), 1, mean, na.rm = TRUE))
    obs <- data.frame(x = coredata(mwd1))
    t <- year(index(djf.dse))
    preq1 <- exp(predict(dfit, newdata = q1))
    preq1[preq1 > 90] <- NA
    tr1 <- predict(lm(preq1 ~ t + I(t^2) + I(t^3) + I(t^4) + 
        I(t^5)))
    tr1[!is.finite(preq1)] <- NA
    preq2 <- exp(predict(dfit, newdata = q2))
    preq2[preq2 > 90] <- NA
    tr2 <- predict(lm(preq2 ~ t + I(t^2) + I(t^3) + I(t^4) + 
        I(t^5)))
    tr2[!is.finite(preq2)] <- NA
    prem <- exp(predict(dfit, newdata = qm))
    prem[prem > 90] <- NA
    tr3 <- predict(lm(prem ~ t + I(t^2) + I(t^3) + I(t^4) + I(t^5)))
    tr3[!is.finite(prem)] <- NA
    Nwd <- zoo(cbind(preq1, preq2, prem, tr1, tr2, tr3), order.by = t)
    nwd.pre <- zoo(exp(predict(dfit, newdata = obs)), order.by = year(mwd1))
    if (plot) {
        if (new) 
            dev.new()
        par(bty = "n")
        plot(zoo(djf.dse, order.by = year(djf.dse)), plot.type = "single", 
            col = rgb(0.5, 0.5, 0.5, 0.2), ylab = expression(paste("mean temperature", 
                (degree * C))), xlab = "", main = loc(x))
        points(mwd1, pch = 19)
        grid()
        if (new) 
            dev.new()
        par(bty = "n")
        plot(Nwd, plot.type = "single", lwd = 5, main = loc(x), 
            ylim = c(0, 90), xlab = "", ylab = paste("number of hot days: T(2m) > ", 
                threshold, unit(x)), col = c(rgb(0.5, 0.5, 0.7, 
                0.5), rgb(0.8, 0.5, 0.5, 0.5), rgb(0.8, 0.5, 
                0.8, 0.5), rgb(0.3, 0.3, 0.6, 0.5), rgb(0.6, 
                0.3, 0.3, 0.5), rgb(0.6, 0.3, 0.6, 0.5)), ...)
        grid()
        points(nwd1, pch = 19)
        lines(nwd.pre, col = rgb(0.5, 0.5, 0.5, 0.5))
    }
    Nwd <- attrcp(x, Nwd)
    attr(Nwd, "unit") <- "days"
    attr(Nwd, "info") <- paste("number of hot days: t2m > ", 
        threshold)
    attr(Nwd, "observation") <- nwd1
    attr(Nwd, "nwd.pre") <- nwd.pre
    index(Nwd) <- t
    class(Nwd) <- c("nevents", "zoo")
    invisible(Nwd)
}
```

@@

  points(1-x$sd.ratio,rep(i,n),pch=pch,col=col,cex=cex)
  text(xlim[1],i,gcmnm.i,pos=4,cex=0.7)
}

biasdiag <- function(predictor,param='slp',
                    path="CMIP5.monthly/",pattern="slp_Amon_ens_",type='ncdf',
                    period=c(1950,2015),rcp='rcp45',verbose=TRUE,
                    lon=c(0,100),lat=c(65,90),select=NULL,xlim=NULL) {

@@

  it <- month.abb[as.numeric(rownames(table(month(predictor))))]
  if (verbose) print(paste('biasdiag, it=',it))

@@

  d.y <- 6 # Number of leading EOFs to evalue
  years <- 1900:2100
  m <- length(years)
  months <- rep(month(y)[1],m)
  X <- matrix(rep(NA,N*m*d.y),N,m*d.y)
  dim(X) <- c(d.y[2],N,m)
  gcmnm <- rep("",N)
  scorestats <- matrix(rep(NA,N*9),N,9)
  colnames(scorestats) <- c("1-r.xval","mean.diff","sd.ratio","autocorr.ratio",
                            "res.trend","res.K-S","res.ar1",'amplitude.ration',
                            "1-R2")

@@

  ## Set up a list environment to keep all the results
  results <- list() 
  if (verbose) print("loop...") 
  for (i in 1:N) {
    if (verbose) print(ncfiles[select[i]])
    gcm <- retrieve(ncfile = ncfiles[select[i]],type=type,
                          lon=range(lon(predictor)),
                          lat=range(lat(predictor)),verbose=verbose)
    if (!is.null(it)) {
      if (verbose) print('Extract some months ot a time period')
      if (verbose) print(it)
      gcm <- subset(gcm,it=it)
    }
    #gcmnm[i] <- attr(gcm,'model_id')
    gcmnm.i <- paste(attr(gcm,'model_id'),attr(gcm,'realization'),sep="-r")
    gcmnm.i <- gsub(' ','',gcmnm.i)
    gcmnm.i <- gsub('-','.',gcmnm.i)
    print(gcmnm.i)
    GCM <- aggregate(subset(gcm,it=month.abb[month(y)[1]]),year,FUN='mean') 
    SLPGCM <- combine(predictor,GCM)
    Z <- try(EOF(SLPGCM))
    diag <- diagnose(Z)
    showceofvar(diag,N,i,gcmnm.i,add=(i!=1),xlim=xlim,verbose=verbose)
    eval(parse(text=paste('results$',it,'.',gcmnm.i,' <- diag',sep='')))
  }
  return(results)
}

@@

```{r fig.width=6, fig.height=6}
ndhr.cal <- data.frame(y = coredata(subset(subset(ndhr100,is=2),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr.cal,family="poisson")
plot(1949:2004,ndhr.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr)[2])
grid()
lines(1949:2004,exp(predict(ds.ndhr)),col='red')
```

@@

```{r fig.width=6, fig.height=6}
## Delhi
ndhr50.1.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=1),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.1 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.1.cal,family="poisson")
plot(1949:2004,ndhr50.1.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr)[1])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.1)),col='red')

# Bangalore
ndhr50.3.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=3),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.3 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.3.cal,family="poisson")
plot(1949:2004,ndhr50.3.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr)[3])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.3)),col='red')
```

@@

```{r}
print(paste('Mean number=',mean(ndhr[,2],na.rm=TRUE),'Variance=',var(ndhr[,2],na.rm=TRUE),
            'should equal',mean(ndhr[,2],na.rm=TRUE),'for Poisson process'))
h <- hist(coredata(ndhr[,2]),freq=FALSE,col='grey',
     main=loc(ndhr)[2])
lines(h$mids-0.5,dpois(x=h$mids-0.5,lambda=mean(ndhr[,2],na.rm=TRUE)),lwd=2,col='red')
grid()
```

@@

```{r}
print('Compare common EOFs')
diag.slp <- biasdiag(slp)
```

>>>>

```{r,echo=TRUE}
library(esd)
library(ncdf)
setwd('~/R')
#rm(list=ls())
## Information about the system and session
```

@@

```{r, eval=FALSE}
hotsummerdays <- function (x, y = NULL, dse = NULL, it = "jja", threshold = 30, 
    verbose = FALSE, plot = TRUE, nmin = 90, new = TRUE, ...) 
{
    if (verbose) 
        print("mildwinterdays")
    stopifnot(inherits(x, "station"))
    if (is.null(y)) 
        y <- x
    djf <- subset(x, it = it)
    djfy <- subset(y, it = it)
    nwd1 <- annual(djfy, FUN = "count", threshold = threshold, 
        nmin = nmin)
    mwd1 <- annual(djf, FUN = "mean", nmin = nmin)
    cal <- data.frame(x = c(coredata(mwd1)), y = c(coredata(nwd1)))
    dfit <- glm(y ~ x, family = "poisson", data = cal)
    if (plot) {
        par(bty = "n")
        plot(cal, pch = 19, ylim = c(0, 90), xlab = expression(paste("mean temperature ", 
            (degree * C))), ylab = "number of hot days", main = loc(x))
        pre <- data.frame(x = seq(min(cal$x, na.rm = TRUE) - 
            1, max(cal$x, na.rm = TRUE) + 5, by = 0.1))
        lines(pre$x, exp(predict(dfit, newdata = pre)), col = rgb(1, 
            0, 0, 0.3), lwd = 3)
        djf.sd <- sd(coredata(djf), na.rm = TRUE)
        qqnorm(coredata(djf))
        qqline(coredata(coredata(djf)), col = "red")
        grid()
    }
    if (is.null(dse)) 
        dse <- DSensemble.t2m(x, biascorrect = TRUE, verbose = verbose, 
            plot = plot)
    djf.dse <- subset(dse, it = "djf")
    index(djf.dse) <- year(djf.dse)
    ovl <- window(djf.dse, start = year(start(x)), end = year(end(x)))
    djf.dse <- djf.dse - mean(coredata(ovl), na.rm = TRUE) + 
        mean(coredata(mwd1), na.rm = TRUE)
    q1 <- data.frame(x = apply(coredata(djf.dse), 1, quantile, 
        probs = 0.05, na.rm = TRUE))
    q2 <- data.frame(x = apply(coredata(djf.dse), 1, quantile, 
        probs = 0.95, na.rm = TRUE))
    qm <- data.frame(x = apply(coredata(djf.dse), 1, mean, na.rm = TRUE))
    obs <- data.frame(x = coredata(mwd1))
    t <- year(index(djf.dse))
    preq1 <- exp(predict(dfit, newdata = q1))
    preq1[preq1 > 90] <- NA
    tr1 <- predict(lm(preq1 ~ t + I(t^2) + I(t^3) + I(t^4) + 
        I(t^5)))
    tr1[!is.finite(preq1)] <- NA
    preq2 <- exp(predict(dfit, newdata = q2))
    preq2[preq2 > 90] <- NA
    tr2 <- predict(lm(preq2 ~ t + I(t^2) + I(t^3) + I(t^4) + 
        I(t^5)))
    tr2[!is.finite(preq2)] <- NA
    prem <- exp(predict(dfit, newdata = qm))
    prem[prem > 90] <- NA
    tr3 <- predict(lm(prem ~ t + I(t^2) + I(t^3) + I(t^4) + I(t^5)))
    tr3[!is.finite(prem)] <- NA
    Nwd <- zoo(cbind(preq1, preq2, prem, tr1, tr2, tr3), order.by = t)
    nwd.pre <- zoo(exp(predict(dfit, newdata = obs)), order.by = year(mwd1))
    if (plot) {
        par(bty = "n")
        plot(zoo(djf.dse, order.by = year(djf.dse)), plot.type = "single", 
            col = rgb(0.5, 0.5, 0.5, 0.2), ylab = expression(paste("mean temperature", 
                (degree * C))), xlab = "", main = loc(x))
        points(mwd1, pch = 19)
        grid()
        par(bty = "n")
        plot(Nwd, plot.type = "single", lwd = 5, main = loc(x), 
            ylim = c(0, 90), xlab = "", ylab = paste("number of hot days: T(2m) > ", 
                threshold, unit(x)), col = c(rgb(0.5, 0.5, 0.7, 
                0.5), rgb(0.8, 0.5, 0.5, 0.5), rgb(0.8, 0.5, 
                0.8, 0.5), rgb(0.3, 0.3, 0.6, 0.5), rgb(0.6, 
                0.3, 0.3, 0.5), rgb(0.6, 0.3, 0.6, 0.5)), ...)
        grid()
        points(nwd1, pch = 19)
        lines(nwd.pre, col = rgb(0.5, 0.5, 0.5, 0.5))
    }
    Nwd <- attrcp(x, Nwd)
    attr(Nwd, "unit") <- "days"
    attr(Nwd, "info") <- paste("number of hot days: t2m > ", 
        threshold)
    attr(Nwd, "observation") <- nwd1
    attr(Nwd, "nwd.pre") <- nwd.pre
    index(Nwd) <- t
    class(Nwd) <- c("nevents", "zoo")
    invisible(Nwd)
}
```

@@

  points(1-x$sd.ratio,rep(i,n),pch=pch,col=col,cex=cex)
  #print(paste(' > ',i,i,gcmnm.i[i]))
  text(xlim[1],i,gcmnm.i[i],pos=4,cex=0.6)
}

biasdiag <- function(predictor,param='slp',it=NULL,
                    path="CMIP5.monthly/",pattern="psl_Amon_ens_",type='ncdf4',
                    period=c(1950,2015),rcp='rcp45',verbose=TRUE,
                    lon=c(0,100),lat=c(65,90),select=NULL,xlim=NULL) {

@@

  if (verbose) print(paste('biasdiag, it=',it))

@@

  nceofs <- 6 # Number of leading EOFs to evalue
  years <- 1900:2100
  m <- length(years)
  months <- rep(month(nceofs),m)
  X <- matrix(rep(NA,N*m*nceofs),N,m*nceofs)
  dim(X) <- c(nceofs,N,m)
  gcmnm.i <- rep("",N)
  scorestats <- matrix(rep(NA,N*9),N,9)
  colnames(scorestats) <- c("1-r.xval","mean.diff","sd.ratio","autocorr.ratio",
                            "res.trend","res.K-S","res.ar1",'amplitude.ration',
                            "1-R2")

@@

  ## Set up a list environment to keep all the results
  results <- list() 
  if (!file.exists('diag.ceof.slp.rda')) {
    if (verbose) print("loop...") 
    for (i in 1:N) {

      if (verbose) print(ncfiles[select[i]])
      gcm <- retrieve(ncfile = ncfiles[select[i]],type=type,
                            lon=range(lon(predictor)),
                           lat=range(lat(predictor)),verbose=verbose)
      if (!is.null(it)) {
        if (verbose) print('Extract some months ot a time period')
        if (verbose) print(it)
        gcm <- subset(gcm,it=it)
        gcm <- subset(gcm,it=range(year(predictor)))
      }
      gcmnm <- paste(attr(gcm,'model_id'),attr(gcm,'realization'),sep="-r")
      gcmnm <- gsub(' ','',gcmnm)
      gcmnm <- gsub('-','.',gcmnm)
      gcmnm.i[i] <- gcmnm
      print(gcmnm.i[i])
      GCM <- aggregate(gcm,year,FUN='mean') 
      SLPGCM <- combine(predictor,GCM)
      Z <- try(EOF(SLPGCM))
      diag <- diagnose(Z)
      showceofvar(diag,N,i,gcmnm.i,add=(i!=1),xlim=xlim,verbose=verbose)
      eval(parse(text=paste('results$member.',i,'.',gcmnm.i[i],' <- diag',sep='')))
    } 
    save(file='diag.ceof.slp.rda',results,gcmnm.i)
  } else {
    load('diag.ceof.slp.rda')
    N <- length(gcmnm.i)
    for (i in 1:N) {
      #print(paste(' < ', i,gcmnm.i[i]))
      showceofvar(results[[i]],N,i,gcmnm.i,add=(i!=1),xlim=xlim,verbose=verbose)
    }
  }
  dev.copy2pdf(file=paste('diag.ceof.slp.india.pdf',sep=''))
  return(results)
}

@@

```{r fig.width=6, fig.height=6}
ndhr.cal <- data.frame(y = coredata(subset(subset(ndhr100,is=2),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr.cal,family="poisson")
plot(1949:2004,ndhr.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr100)[2])
grid()
lines(1949:2004,exp(predict(ds.ndhr)),col='red')
```

@@

```{r fig.width=6, fig.height=6}
## Delhi
ndhr50.1.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=1),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.1 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.1.cal,family="poisson")
plot(1949:2004,ndhr50.1.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr50)[1])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.1)),col='red')

# Bangalore
ndhr50.3.cal <- data.frame(y = coredata(subset(subset(ndhr50,is=3),it=c(1949,2004))),
                       X = coredata(subset(eof.es,it=c(1949,2004))))
ds.ndhr50.3 <- glm(y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7, data=ndhr50.3.cal,family="poisson")
plot(1949:2004,ndhr50.3.cal$y,pch=19,ylab=expression(N(X > 100*mm)),xlab='',main=loc(ndhr50)[3])
grid()
lines(1949:2004,exp(predict(ds.ndhr50.3)),col='red')
```

@@

```{r}
print(paste('Mean number=',mean(ndhr100[,2],na.rm=TRUE),'Variance=',var(ndhr100[,2],na.rm=TRUE),
            'should equal',mean(ndhr100[,2],na.rm=TRUE),'for Poisson process'))
h <- hist(coredata(ndhr100[,2]),freq=FALSE,col='grey',
     main=loc(ndhr100)[2])
lines(h$mids-0.5,dpois(x=h$mids-0.5,lambda=mean(ndhr100[,2],na.rm=TRUE)),lwd=2,col='red')
grid()
```

@@

Common EOFs can be used to assess the GCMs' ability to reproduce the spatio-temporal structure in covariance in cinnection to mean sea-level pressure (SLP) variations over the Indian sub-continent. In orther words, whether the GCMs reproduce the observed circulation features such as wind patterns associated with the monsoon. The code below compares the statistical character of the odel results.

```{r}
print('Compare common EOFs')
diag.slp <- biasdiag(slp,it=month.abb[season],verbose=FALSE)
```
